Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-12 _paper.rmd
ames housing needs to have variables explained
-Penguin, Chocolate, and Ames update -Introduce PI/CI, Obs->inst, var->feat

---
title: "Exploring Local Explanations of Non-linear Models Using Animated Linear Projections"
shorttitle: Cheem
keywords:
- R
- XAI
- explanable artificial intelligence
- black-box models
- local explanations
- grand tour
- manual tour
- radial tour
author:
  #- name: "Nicholas Spyrison"
  #   address: |
  #     | ORCiD: 0000-0002-8417-0212
  #   email: \email{nicholas.spyrison@monash.edu}
  #   orcid: 0000-0002-8417-0212
  #- name: "Dianne Cook"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 000-0002-3813-7155
  #   orcid: 000-0002-3813-7155
  #- name: "Kimbal Marriott"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 0000-0002-9813-0377
  #   orcid: 0000-0002-9813-0377
abstract: |
 The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models are using predictors, to arrive at a prediction, with a point estimate of the linear variable importance in the vicinity of each observation. These can be considered to be linear projections, and as such can be further explored interactively to better understand the interaction between variables used to make predictions, across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation, and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (football salaries, house prices) response variables. The methods are implemented in the __R__ package __cheem__, available on CRAN.
#documentclass: jmlr
#classoption: article
output:
  ## pdf_document: requires \ref{} 
  #pdf_document:
  ## bookdown::pdf_document2: allows \@ref()
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    number_sections: yes
    fig_caption: yes
    #includes:
    #  in_header: template/jmlr2e.sty
bibliography: cheem_paper.bib
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
  - \usepackage{setspace} ## for title page spacing
  - \usepackage{hyperref} ## for all sorts of linking
---

@@

<!-- Introduce explanatory vs predictive modeling -->
There are different reasons and emphases to fit a model. @breiman_statistical_2001, reiterated by @shmueli_explain_2010, taxonomizes modeling based on its purpose; _explanatory_ modeling is done for some inferential purpose, while _predictive_ modeling focuses more on the predictions of out-of-sample observations. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate non-linear models. The use of black-box models is becoming increasingly common, but not without their share of controversy [@oneil_weapons_2016; @kodiyan_overview_2019]. However, the loss of interpretation presents a challenge.

<!-- Interpretability & baises -->
Interpretability is vital for exploring and protecting against potential biases (e.g. sex [@dastin_amazon_2018; @duffy_apple_2019], race [@larson_how_2016], and age [@diaz_addressing_2018]) in any model. For instance, models regularly pick up on biases in the training data that have observed influence on the response (output) variable, which is then built into the model. Variable-level (feature-level) interpretability of models is essential in the evaluation models for such biases. It is also generally important for many problems, where it is not enough to accurately predict accurately but one must be able to explain which predictors are most responsible in generating a response value.

<!-- Interpretability & data drift -->
Another concern is that of data drift, which is a shift in support or domain of the explanatory variables (feature or predictors). Non-linear models are typically more sensitive, and do not extrapolate well outside of the training data domain. Better interpretability of the model means that there is more transparency where models' predictions may be plausible or completely unreliable.

<!-- Local explanations -->
Explainable Artificial Intelligence (XAI) is an emerging field of research that tries to increase the interpretability of black-box models. A common approach is to use _local explanations_, which attempt to approximate linear variable importance at the location of each observation (instance), or the predictions at a specific point in the data domain. Because these are point specific, a challenge is visualizing them to more comprehensively understand a model.

<!-- Data visualization tours -->
In multivariate data visualization, a _tour_ [@asimov_grand_1985; @buja_grand_1986; @lee_state_2021] is a sequence of linear projections of data onto a lower-dimensional space. Tours are viewed as an animation over minor changes to the projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of that structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the shape of the shadow change conveys information of the structure and features of the object.

<!-- Manual tours --> 
There are various types of tours distinguished by the generation of projection bases. In a _manual_ tour [@cook_manual_1997; @spyrison_spinifex_2020], the path is defined by changing the contribution of a selected variable. <!--tours and models --> Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries [@wickham_visualizing_2015], and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis paths [@lee_pptree_2013; @da_silva_projection_2021].

<!-- Purposed approach -->
In our proposed approach we use the radial manual tour to scrutinize a local explanation. Additional interactivity allows the user to identify an observation of interest, then explore its local explanation by changing variable contribution with the radial tour. The methods are implemented in R package __cheem__. Example datasets are provided to illustrate usage for classification and regression tasks.

<!-- What-if & counterfactual analysis -->
Using a radial tour can be considered similar to counterfactual, what-if analysis, such as _ceteris paribus_ [@biecek_ceterisparibus_2020]. This phrase, Latin for "other things held constant" or "all else unchanged", shows how an observation's prediction would change from a marginal change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual; it varies contributions of the variables by rotating the basis, a reorientation of the data object. A constraint in our approach is that the basis must remain orthonormal. That means, when the contribution of one variable decreases, the contributions of others necessarily increase such that there is a complete component in that direction. This also ensures that what is seen is strictly a low-dimensional projection from high-dimensions, and is thus an interpretable visualisation.

<!-- Paper structure -->
The remainder of this paper is organized as follows. The following Section, \@ref(sec:explanations) covers the background of the local explanation, and the traditional visuals produced. Section \@ref(sec:tour) explains the animations of continuous linear projections. Section \@ref(sec:applicationdesign) discusses the visual layout in the interactive interface, how they facilitate analysis, data preprocessing, and package infrastructure. Then Section \@ref(#sec:casestudies) illustrates the application to supervised learning with categorical and quantitative response variables. We conclude with Section \@ref(sec:cheemdiscussion) of the insights gained and directions that might be explore in the future.

@@

<!-- Reminder of local explanation -->
Consider a highly non-linear model. It can be hard to determine whether small changes in a variable's value will make a class prediction change group or identify which variables contribute to an extreme residual. Local explanations shed light on these situations by approximating linear variable importance in the vicinity of a single observation or point.

@@

<!-- Uses of local explanations -->
These instance-level explanations are used in various ways depending on the data. In image classification, where pixels would correspond to predictors, saliency maps overlay or offset a heatmap indicating important pixels [@simonyan_deep_2014]. For instance, pixels corresponding to snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words [@vanni_textual_2018]. In the case of numeric regression, they are used to explain variable additive contributions from the model intercept to the observation's prediction [@ribeiro_why_2016].

<!-- SHAP -->
SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation conceptually by examining the effect of other variables on the contribution of the variable of interest on predicting the response. This explanations almost all point to @shapley_value_1953's method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score. @strumbelj_efficient_2010 introduced the use of SHAP for local explanations in ML models. While the models themselves are invariant to the order of the predictors, the additive explanations are not. The SHAP values are the mean contributions over some subset these permutations. This is also related to partial dependence plots [@molnar_interpretable_2020], used to explain the effect of a variable by predicting the response for a range of values on this variable, after fixing the value of all other variables to their mean. Partial dependence plots are a global approximation of the variable importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in @wickham_visualizing_2015, which helps to understand the relative importance of each variable in the context of all other candidate variables.

<!-- tree SHAP -->
For our application, we use _tree SHAP_, a variant of SHAP enjoys a lower computational complexity [@lundberg_consistent_2018]. Instead of aggregating over permutations of the variable order, tree SHAP approximates instance level feature importance by exploring the stucture of the decision trees. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized explanations as the starting projection basis (call this the _attribution projection_) to further scrutinize the explanation.

```{r shapdistrbd, echo=F, fig.align='center', out.width="60%", fig.cap = "Illustration of the use of SHAP values for a random forest model for salaries of FIFA 2020 players based on nine predictors corresponding to different skills. A star offensive and densive player, Messi and a defensive player are compared, L. Messi and V van Dijk respectively. Plot a) is the break-down plot showing one additive variable-attributution. These attributions are sensitive to the order of the variables. Plot b) shows the distribution of attribution for each variable across 25 such permutation of the order of predictors, with the mean displayed as a dot, for each players. Offense and movement are important for Messi but not van Dijk, and conversely defense and power are important for van Dijk but not Messi."}
knitr::include_graphics("./figures/shap_distr_bd.png")
```

<!-- Fifa example -->
Following the use case _Explanatory Model Analysis_ [@biecek_explanatory_2021], we use FIFA data to illustrate the use of SHAP. Consider soccer data from the FIFA 2020 season [@leone_fifa_2020]. There are 5000 observations of 9 skill measures (after aggregating highly correlated variables). A random forest model is fit regressing wages [2020 Euros], from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \@ref(fig:shapdistrbd). We expect to see a difference in the attribution of the variable importance across the two positions of the players, which would be interpreted as how the player's salary depends on this combination of skill sets. Plot (b) is a modified breakdown plot [@gosiewska_ibreakdown_2019] where the order of variables is fixed, so the two observations can be more easily compared.

<!-- Segue -->
In summary, these plots highlight how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, two players with different positions receive different profiles of variable importance to explain the prediction of their wages.

@@

<!-- Tours intro -->
A _tour_ enables viewing of high-dimensional data by animating many linear projections with small incremental changes. It is achieved by following a path of linear projections (bases) of high-dimensional space. One of the features of the tour is the object permanence of the data points; one can track the relative change of observations in time, and as such gain information about the relationships between points across multiple variables. There are various types of tours that are distinguished by how the paths are generated [@lee_state_2021; @cook_grand_2008].

<!-- Manual tour -->
The manual tour [@cook_manual_1997] defines its path by changing a selected variable's contribution to a basis, to allow the variable to contribute more or less to the projection. The contribution of all other variables is constrained by the requirement that a basis needs to be orthonormal (column correspond to vectors, with unit length, and orthogonal to each other).  The manual tour is primarily used to assess the importance of a variable to structure visible in a projection. It also lends itself to pre-computation to be queued in advance or computed on-the-fly for human-in-the-loop analysis [@karwowski_international_2006]. <!-- However, this navigation is relatively time-consuming due to the vast volume of the display space. It is advisable to use this method to explore the sensitivity of the variable contribution to a previously identified feature of interest. In this case, the projection of the normalized explanations is the feature of interest.-->

<!-- Radial tour -->
A version of the manual tour called a _radial tour_ is implemented in @spyrison_spinifex_2020 and forms the basis of the new work. In a radial tour, the selected variable is allowed to change its magnitude of contribution but not its angle; it must move along the direction of its original contribution. The implementation allows for pre-computation and also interactive re-calculation to focus on a different variable.

@@

To explore the local explanations, an ensemble of plots [@unwin] is provided, called the _cheem viewer_. There are two primary plots: the global view to give the context of all of the SHAP values, and the radial tour view to explore the local explanations with user-controlled rotation. In addition, there are numerous user inputs, including variable selection for the radial tour, and instance selection for making comparisons. Figures \@ref(fig:classificationcase) and \@ref(fig:regressioncase) contain screenshots showing the cheem viewer for the two primary tasks classification (categorical response) and regression (quantitative response).

@@

<!-- Purpose -->
The global view provides the context of all observations and facilitates the exploration of the separability of the data- and attribution-spaces. Both of these spaces are of dimension $n\times p$, where $n$ is the number of observations and $p$ is the number of predictors. The attribution space corresponds to the local explanations for each observation, which will have $p$ values for each observation.

<!-- Approximations of the spaces, PC1:2 -->
A visualisation of these spaces is provided by the first two principal components of their respective spaces. In addition, a plot observed by predicted response is also provided. In both PCA plots the orientation and magnitude of the variables are inscribed on a unit circle, similar to a biplot [@gabriel_biplot_1971]. A single 2D projection will not encompass all of the structure of higher-dimensional space, but it is generally a useful visual summary. <!-- Interactions --> For classification tasks, misclassified observations are circled in red if applicable. Linked brushing between the plots is provided and a tabular display of selected points helps to facilitate exploration of the spaces and the model.

While the comparison of these spaces is interesting, a main purpose of the global view is to enable the selection of observations, from which to explore the local explanations.

@@

<!-- A function of the obs selected -->
The global view facilitated the selection of a primary and optional comparison observation. The variable-level attribution of the primary observation is normalized and used as the initial 1D basis in a radial tour. This is an approximation of the contributions of the linear variables that best explain the difference between the model intercept and an observation's prediction, not the local shape of the model surface.

<!-- Tour start frame -->
The initial frame is the normalized SHAP values of the primary observation. The current projection basis is depicted as the width of a bar, the variable's contribution to the horizontal axis. The normalized values of all observations are shown as vertical parallel coordinate plots.

<!-- Tour animation -->
The radial tour creates a basis path by varying the contribution of a selected variable, fully into and out of a projection frame. Doing so tests an individual variable's sensitivity to the structure identified by the local explanation. The default variable selected has the largest discrepancy between the attribution of primary and comparison observations. The following sections elaborate on the takeaways from applying this approach in classification and regression tasks. <!-- Segue to classification -->Now that we have introduced the global view and corresponding cheem radial tour, let us discuss the differences between the classification and regression cases.

@@

Typically we select a misclassified observation compared to a correctly classified point nearby in data space. The model infomation in the global view is a depiction of the model conmfusion matrix. The radial tour is 1D, with density display, while the goal of the tour; exploring the sensitivity of each variable to structure identified by the local explanation, evaluating the support or robustness of the prediction.

```{r classificationcase, echo=F, out.width = "100%", fig.align="center", fig.cap = "Overview of the cheem viewer for classification tasks. Plots are colored on predicted class, and red circles indicate misclassified observations. Model visual is essentially a representation of the confusion matrix of the model. The radial tour is a 1D projection starting the attribution projection. The first frame is the linear variable importances that best describe the observation's prediction. We probe the support of variable contributions by selecting a variable and use the radial tour to vary its coefficient."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

In the regression case, the global view can be colored on a statistic to highlight the explanation space's structure. For this purpose, we include residuals, log Mahalanobis distance of data space (a measure of outlyingness), and the correlation of the attribution projection with the observed response. In the radial tour, the horizontal positions are the same, the basis projection of the radial tour. The vertical position is fixed to the observed response variable and residuals in the middle and right panels. Correspondingly, the display changes from univariate density to 2D scatterplot. The basis is still one component (horizontal) independent of the vertical position.

@@

<!-- Exploratory interactions -->
A tooltip displays observation number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots. The information corresponding to the selected points is populated on a dynamic table. These interactions aid exploration of the spaces and, finally, identification of a primary and comparison observation.

```{r interactions, echo=F, out.width="100%", fig.cap = "Illustration of data explorations interactions in the global view. This view has linked brushing, where observations selected in one facet are highlighted in the other facets and populate an interactive tabular display below. Tooltips display when hovering over an observation."}
knitr::include_graphics("./figures/app_interactions.PNG")
```

@@

(ref:citeRf) @liaw_classification_2002
(ref:citeTs) @kominsarczyk_treeshap_2021
\begin{itemize}
	\item \textbf{Data:} predictors and response are unscaled complete numerical matrix. Most models and local explanations are scale invariant.
	\item \textbf{Model and explanation:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [(ref:citeRf)], compatibility tree SHAP. We use modest hyperparameters, namely: 125 trees, number variables randomly sampled at each split, mtry = $\sqrt{p}$ or $p/3$ for classification and regression, and minimum size of terminal nodes $max(1, n/500)$ or $max(5, n/500)$ for classification and regression. Tree SHAP is calculated for \emph{each} observation using the package \textbf{treeshap} (ref:citeTs). This implementation aggregates over exhaustively over the attribution of all trees and we opt to not to fit interactions of features. 
	\item \textbf{Cheem view:} after the model and full explanation space are calculated we scale each feature by standard deviations away from the mean to achieve a common support for visuals. Statistics for mapping to color are calculated on the scaled spaces. Interactive tabular display reports the original values.
\end{itemize}

>>>>

---
title: "Exploring Local Explanations of Non-linear Models Using Animated Linear Projections"
shorttitle: Cheem
keywords:
- R
- XAI
- explanable artificial intelligence
- black-box models
- local explanations
- grand tour
- manual tour
- radial tour
author:
  #- name: "Nicholas Spyrison"
  #   address: |
  #     | ORCiD: 0000-0002-8417-0212
  #   email: \email{nicholas.spyrison@monash.edu}
  #   orcid: 0000-0002-8417-0212
  #- name: "Dianne Cook"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 000-0002-3813-7155
  #   orcid: 000-0002-3813-7155
  #- name: "Kimbal Marriott"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 0000-0002-9813-0377
  #   orcid: 0000-0002-9813-0377
abstract: |
 The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models are using predictors, to arrive at a prediction, with a point estimate of the linear feature importance in the vicinity of each instance. These can be considered to be linear projections, and as such can be further explored interactively to better understand the interaction between features used to make predictions, across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any instance, and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (football salaries, house prices) response features. The methods are implemented in the __R__ package __cheem__, available on CRAN.
#documentclass: jmlr
#classoption: article
output:
  ## pdf_document: requires \ref{} 
  #pdf_document:
  ## bookdown::pdf_document2: allows \@ref()
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    number_sections: yes
    fig_caption: yes
    #includes:
    #  in_header: template/jmlr2e.sty
bibliography: cheem_paper.bib
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
  - \usepackage{setspace} ## for title page spacing
  - \usepackage{hyperref} ## for all sorts of linking
---

@@

<!-- Introduce explanatory vs predictive modeling -->
There are different reasons and emphases to fit a model. @breiman_statistical_2001, reiterated by @shmueli_explain_2010, taxonomizes modeling based on its purpose; _explanatory_ modeling is done for some inferential purpose, while _predictive_ modeling focuses more on the predictions of out-of-sample instances. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate non-linear models. The use of black-box models is becoming increasingly common, but not without their share of controversy [@oneil_weapons_2016; @kodiyan_overview_2019]. However, the loss of interpretation presents a challenge.

<!-- Interpretability & baises -->
Interpretability is vital for exploring and protecting against potential biases (e.g. sex [@dastin_amazon_2018; @duffy_apple_2019], race [@larson_how_2016], and age [@diaz_addressing_2018]) in any model. For instance, models regularly pick up on biases in the training data that have observed influence on the response (output) feature, which is then built into the model. Feature-level (variable-level) interpretability of models is essential in the evaluation models for such biases. It is also generally important for many problems, where it is not enough to accurately predict accurately but one must be able to explain which predictors are most responsible in generating a response value.

<!-- Interpretability & data drift -->
Another concern is that of data drift, which is a shift in support or domain of the explanatory features (variable or predictors). Non-linear models are typically more sensitive, and do not extrapolate well outside of the training data domain. Better interpretability of the model means that there is more transparency where models' predictions may be plausible or completely unreliable.

<!-- Local explanations -->
Explainable Artificial Intelligence (XAI) is an emerging field of research that tries to increase the interpretability of black-box models. A common approach is to use _local explanations_, which attempt to approximate linear feature importance at the location of each instance (observation), or the predictions at a specific point in the data domain. Because these are point specific, a challenge is visualizing them to more comprehensively understand a model.

<!-- Data visualization tours -->
In multivariate data visualization, a _tour_ [@asimov_grand_1985; @buja_grand_1986; @lee_state_2021] is a sequence of linear projections of data onto a lower-dimensional space. Tours are viewed as an animation over minor changes to the projection basis. Structure in a projection can then be explored visually to see which features contribute to the formation of that structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the shape of the shadow change conveys information of the structure and features of the object.

<!-- Manual tours --> 
There are various types of tours distinguished by the generation of projection bases. In a _manual_ tour [@cook_manual_1997; @spyrison_spinifex_2020], the path is defined by changing the contribution of a selected feature. <!--tours and models --> Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries [@wickham_visualizing_2015], and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis paths [@lee_pptree_2013; @da_silva_projection_2021].

<!-- Purposed approach -->
In our proposed approach we use the radial manual tour to scrutinize a local explanation. Additional interactivity allows the user to identify an instance of interest, then explore its local explanation by changing feature contribution with the radial tour. The methods are implemented in R package __cheem__. Example datasets are provided to illustrate usage for classification and regression tasks.

<!-- What-if & counterfactual analysis -->
Using a radial tour can be considered similar to counterfactual, what-if analysis, such as _ceteris paribus_ [@biecek_ceterisparibus_2020]. This phrase, Latin for "other things held constant" or "all else unchanged", shows how an instance's prediction would change from a marginal change in one explanatory feature given that other features are held constant. It ignores correlations of the features and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual; it varies contributions of the features by rotating the basis, a reorientation of the data object. A constraint in our approach is that the basis must remain orthonormal. That means, when the contribution of one feature decreases, the contributions of others necessarily increase such that there is a complete component in that direction. This also ensures that what is seen is strictly a low-dimensional projection from high-dimensions, and is thus an interpretable visualisation.

<!-- Paper structure -->
The remainder of this paper is organized as follows. The following Section, \@ref(sec:explanations) covers the background of the local explanation, and the traditional visuals produced. Section \@ref(sec:tour) explains the animations of continuous linear projections. Section \@ref(sec:applicationdesign) discusses the visual layout in the interactive interface, how they facilitate analysis, data preprocessing, and package infrastructure. Then Section \@ref(#sec:casestudies) illustrates the application to supervised learning with categorical and quantitative response features. We conclude with Section \@ref(sec:cheemdiscussion) of the insights gained and directions that might be explore in the future.

@@

<!-- Reminder of local explanation -->
Consider a highly non-linear model. It can be hard to determine whether small changes in a feature's value will make a class prediction change group or identify which features contribute to an extreme residual. Local explanations shed light on these situations by approximating linear feature importance in the vicinity of a single instance.

@@

<!-- Uses of local explanations -->
These instance-level explanations are used in various ways depending on the data. In image classification, where pixels would correspond to predictors, saliency maps overlay or offset a heatmap indicating important pixels [@simonyan_deep_2014]. For instance, pixels corresponding to snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words [@vanni_textual_2018]. In the case of numeric regression, they are used to explain feature additive contributions from the model intercept to the instance's prediction [@ribeiro_why_2016].

<!-- SHAP -->
SHaply Additive exPlanations (SHAP) approximates the feature importance in the vicinity of one instance conceptually by examining the effect of other features on the contribution of the feature of interest on predicting the response. This explanations almost all point to @shapley_value_1953's method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score. @strumbelj_efficient_2010 introduced the use of SHAP for local explanations in ML models. While the models themselves are invariant to the order of the predictors, the additive explanations are not. The SHAP values are the mean contributions over some subset these permutations. This is also related to partial dependence plots [@molnar_interpretable_2020], used to explain the effect of a feature by predicting the response for a range of values on this feature, after fixing the value of all other features to their mean. Partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in @wickham_visualizing_2015, which helps to understand the relative importance of each feature in the context of all other candidate features.

<!-- tree SHAP -->
For our application, we use _tree SHAP_, a variant of SHAP enjoys a lower computational complexity [@lundberg_consistent_2018]. Instead of aggregating over permutations of the feature order, tree SHAP approximates instance level feature importance by exploring the stucture of the decision trees. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized explanations as the starting projection basis (call this the _attribution projection_) to further scrutinize the explanation.

```{r shapdistrbd, echo=F, fig.align='center', out.width="60%", fig.cap = "Illustration of the use of SHAP values for a random forest model for salaries of FIFA 2020 players based on nine predictors corresponding to different skills. A star offensive and densive player, Messi and a defensive player are compared, L. Messi and V van Dijk respectively. Plot a) is the break-down plot showing one additive feature-attributution. These attributions are sensitive to the order of the features. Plot b) shows the distribution of attribution for each feature across 25 such permutation of the order of predictors, with the mean displayed as a dot, for each players. Offense and movement are important for Messi but not van Dijk, and conversely defense and power are important for van Dijk but not Messi."}
knitr::include_graphics("./figures/shap_distr_bd.png")
```

<!-- Fifa example -->
Following the use case _Explanatory Model Analysis_ [@biecek_explanatory_2021], we use FIFA data to illustrate the use of SHAP. Consider soccer data from the FIFA 2020 season [@leone_fifa_2020]. There are 5000 instances of 9 skill measures (after aggregating highly correlated features). A random forest model is fit regressing wages [2020 Euros], from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \@ref(fig:shapdistrbd). We expect to see a difference in the attribution of the feature importance across the two positions of the players, which would be interpreted as how the player's salary depends on this combination of skill sets. Plot (b) is a modified breakdown plot [@gosiewska_ibreakdown_2019] where the order of features is fixed, so the two instances can be more easily compared.

<!-- Segue -->
In summary, these plots highlight how local explanations bring interpretability to a model, at least in the vicinity of their instances. In this instance, two players with different positions receive different profiles of feature importance to explain the prediction of their wages.

@@

<!-- Tours intro -->
A _tour_ enables viewing of high-dimensional data by animating many linear projections with small incremental changes. It is achieved by following a path of linear projections (bases) of high-dimensional space. One of the features of the tour is the object permanence of the data points; one can track the relative change of instances in time, and as such gain information about the relationships between points across multiple features. There are various types of tours that are distinguished by how the paths are generated [@lee_state_2021; @cook_grand_2008].

<!-- Manual tour -->
The manual tour [@cook_manual_1997] defines its path by changing a selected feature's contribution to a basis, to allow the feature to contribute more or less to the projection. The contribution of all other features is constrained by the requirement that a basis needs to be orthonormal (column correspond to vectors, with unit length, and orthogonal to each other).  The manual tour is primarily used to assess the importance of a feature to structure visible in a projection. It also lends itself to pre-computation to be queued in advance or computed on-the-fly for human-in-the-loop analysis [@karwowski_international_2006]. <!-- However, this navigation is relatively time-consuming due to the vast volume of the display space. It is advisable to use this method to explore the sensitivity of the feature contribution to a previously identified feature of interest. In this case, the projection of the normalized explanations is the feature of interest.-->

<!-- Radial tour -->
A version of the manual tour called a _radial tour_ is implemented in @spyrison_spinifex_2020 and forms the basis of the new work. In a radial tour, the selected feature is allowed to change its magnitude of contribution but not its angle; it must move along the direction of its original contribution. The implementation allows for pre-computation and also interactive re-calculation to focus on a different feature.

@@

To explore the local explanations, an ensemble of plots [@unwin] is provided, called the _cheem viewer_. There are two primary plots: the global view to give the context of all of the SHAP values, and the radial tour view to explore the local explanations with user-controlled rotation. In addition, there are numerous user inputs, including feature selection for the radial tour, and instance selection for making comparisons. Figures \@ref(fig:classificationcase) and \@ref(fig:regressioncase) contain screenshots showing the cheem viewer for the two primary tasks: classification (categorical response) and regression (quantitative response).

@@

<!-- Purpose -->
The global view provides the context of all instances and facilitates the exploration of the separability of the data- and attribution-spaces. Both of these spaces are of dimension $n\times p$, where $n$ is the number of instances and $p$ is the number of predictors. The attribution space corresponds to the local explanations for each instance, which will have $p$ values for each instance.

<!-- Approximations of the spaces, PC1:2 -->
A visualisation of these spaces is provided by the first two principal components of their respective spaces. In addition, a plot observed by predicted response is also provided. In both PCA plots the orientation and magnitude of the features are inscribed on a unit circle, similar to a biplot [@gabriel_biplot_1971]. A single 2D projection will not encompass all of the structure of higher-dimensional space, but it is generally a useful visual summary. <!-- Interactions --> For classification tasks, misclassified instances are circled in red if applicable. Linked brushing between the plots is provided and a tabular display of selected points helps to facilitate exploration of the spaces and the model.

While the comparison of these spaces is interesting, a main purpose of the global view is to enable the selection of instances, from which to explore the local explanations. The projection attribution of the Primary Instance (PI) is explored and typically viewed with an optional Comparison Instance (CI).

@@

<!-- A function of the obs selected -->
The global view facilitated the selection instances. The feature-level attribution of the primary instance is normalized and used as the initial 1D basis in a radial tour. This is an approximation of the contributions of the linear features that best explain the difference between the model intercept and an instance's prediction, not the local shape of the model surface.

<!-- Tour start frame -->
The initial frame is the normalized SHAP values of the PI. The current projection basis is depicted as the width of a bar, the feature's contribution to the horizontal axis. The normalized values of all instances are shown as vertical parallel coordinate plots.

<!-- Tour animation -->
The radial tour creates a basis path by varying the contribution of a selected feature, fully into and out of a projection frame. Doing so tests an individual feature's sensitivity to the structure identified by the local explanation. The default feature selected has the largest discrepancy between the attribution of primary and comparison instances. The following sections elaborate on the takeaways from applying this approach in classification and regression tasks. <!-- Segue to classification -->Now that we have introduced the global view and corresponding cheem radial tour, let us discuss the differences between the classification and regression cases.

@@

Typically we select a misclassified instance compared to a correctly classified point nearby in data space. The model infomation in the global view is a depiction of the model conmfusion matrix. The radial tour is 1D, with density display, while the goal of the tour; exploring the sensitivity of each feature to structure identified by the local explanation, evaluating the support or robustness of the prediction.

```{r classificationcase, echo=F, out.width = "100%", fig.align="center", fig.cap = "Overview of the cheem viewer for classification tasks. Plots are colored on predicted class, and red circles indicate misclassified instances. Model visual is essentially a representation of the confusion matrix of the model. The radial tour is a 1D projection starting the attribution projection. The first frame is the linear feature importances that best describe the instance's prediction. We probe the support of feature contributions by selecting a feature and use the radial tour to vary its coefficient."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

In the regression case, the global view can be colored on a statistic to highlight the explanation space's structure. For this purpose, we include residuals, log Mahalanobis distance of data space (a measure of outlyingness), and the correlation of the attribution projection with the observed response. In the radial tour, the horizontal positions are the same, the basis projection of the radial tour. The vertical position is fixed to the observed response feature and residuals in the middle and right panels. Correspondingly, the display changes from univariate density to 2D scatterplot. The basis is still one component (horizontal) independent of the vertical position.

@@

<!-- Exploratory interactions -->
A tooltip displays instance number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots. The information corresponding to the selected points is populated on a dynamic table. These interactions aid exploration of the spaces and, finally, identification of a primary and comparison instance.

```{r interactions, echo=F, out.width="100%", fig.cap = "Illustration of data explorations interactions in the global view. This view has linked brushing, where instances selected in one facet are highlighted in the other facets and populate an interactive tabular display below. Tooltips display when hovering over an instance."}
knitr::include_graphics("./figures/app_interactions.PNG")
```

@@

(ref:citeRf) @liaw_classification_2002
(ref:citeTs) @kominsarczyk_treeshap_2021
\begin{itemize}
	\item \textbf{Data:} predictors and response are unscaled complete numerical matrix. Most models and local explanations are scale invariant.
	\item \textbf{Model and explanation:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [(ref:citeRf)], compatibility tree SHAP. We use modest hyperparameters, namely: 125 trees, number features randomly sampled at each split, mtry = $\sqrt{p}$ or $p/3$ for classification and regression, and minimum size of terminal nodes $max(1, n/500)$ or $max(5, n/500)$ for classification and regression. Tree SHAP is calculated for \emph{each} instance using the package \textbf{treeshap} (ref:citeTs). This implementation aggregates over exhaustively over the attribution of all trees and we opt to not to fit interactions of features. 
	\item \textbf{Cheem view:} after the model and full explanation space are calculated we scale each feature by standard deviations away from the mean to achieve a common support for visuals. Statistics for mapping to color are calculated on the scaled spaces. Interactive tabular display reports the original values.
\end{itemize}

<!-- Note on time of execution -->
The time to preprocess the data will vary significantly with the model and local explanation. For reference, the FIFA data, 5000 instances of nine explanatory features, took 2.5 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each instance took 270 seconds combined. PCA and statistics of the features and attributions took 2.8 seconds. These runtimes were from a non-parallelized R session on a modern laptop, but suffice to say that the bulk of the time will be spent on the local attribution. An increase in model complexity or data dimensionality will quickly become an obstacle. With its reduced computational complexity, this makes tree SHAP a good candidate to start with. Alternatively, the package __fastshap__ [@greenwell_fastshap_2020] claims extremely low runtimes, which are attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.
